Automatically Splitting a Two-Stage Lambda Calculus

### 2015-07-11 paper/abstract.tex
various small edits

Staged programming languages assign a stage to each program expression
and evaluate each expression in its assigned stage. A common use of
staged languages is to describe programs where inputs arrive at
different times or rates. In this paper we present an algorithm for
statically {\em splitting} these mixed-staged programs into two
unstaged, but dependent, programs where the outputs of the first
program can be efficiently reused across multiple invocations of the
second. While previous algorithms for performing this transformation
(also called \emph{pass separation} and \emph{data specialization})
were limited to operate on simpler, imperative languages, we define a
splitting algorithm for an explicitly-two-stage, typed lambda
calculus \lang\ with a $\fut$ modality denoting computation at a later
stage, and a $\curr$ modality noting purely first stage code.  Most
notably, the algorithm splits mixed-stage recursive and higher-order
functions.  We have proven the dynamic correctness of our splitting
algorithm with respect to a partial-evaluation semantics, and
mechanized this proof in Twelf. We have also implemented the algorithm
in a prototype compiler, and demonstrate that the ability to split
programs in a language featuring recursion and higher-order features
enables non-trivial algorithmic transformations that asymptotically
improve code efficiency by nearly a linear factor, and also
facilitates modular expression of staged programs.

>>>>

Staged programming languages assign a stage to each program expression
and evaluate each expression in its assigned stage. A common use of
staged languages is to describe programs where inputs arrive at
different times or rates. In this paper we present an algorithm for
statically {\em splitting} these mixed-staged programs into two
unstaged, but dependent, programs where the outputs of the first
program can be efficiently reused across multiple invocations of the
second. While previous algorithms for performing this transformation
(also called \emph{pass separation} and \emph{data specialization})
were limited to operate on simpler, imperative languages, we define a
splitting algorithm for an explicitly-two-stage, typed lambda
calculus \lang\ with a $\fut$ modality denoting computation at a later
stage, and a $\curr$ modality noting purely first stage code.  Most
notably, the algorithm splits mixed-stage recursive and higher-order
functions.  We prove the dynamic correctness of our splitting
algorithm with respect to a partial-evaluation semantics, and
mechanize this proof in Twelf. We also implement the algorithm
in a prototype compiler, and demonstrate that the ability to split
programs in a language featuring recursion and higher-order features
enables non-trivial algorithmic transformations that asymptotically
improve code efficiency by nearly a linear factor and also
facilitates modular expression of staged programs.

### 2015-07-11 paper/intro.tex
figure fiddling
various small edits

Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_1, y_2 \ldots, y_m)$@ = 
  @$f(x, y_1);  f(x, y_2); \ldots; f(x, y_m)$@
\end{lstlisting}
%
In this scenario, a multivariate function is applied many times with
one argument fixed, so it is advantageous to find the computations in
$f$ that depend on $x$ and then stage program execution so they are
performed only once at the beginning of execution.

@@

J{\o}rring and Scherlis classify automatic program staging
transformations, such as the one described above, as forms of {\em
  frequency reduction} or {\em
  precomputation}\,\cite{JS86-staging}. To perform frequency
reduction, one identifies and hoists computations that are performed
multiple times, in order to compute them only once. To perform
precomputation, one identifies computations that can be performed in
advance, and does so---for example, at compile time if the relevant
inputs are statically known.

@@

While simple forms of frequency reduction include standard compiler
optimizations such as loop hoisting and common subexpression
elimination, J{\o}rring and Scherlis proposed the more general
transformation of {\em splitting} a program into multiple subfunctions
(called pass separation in \,\cite{JS86-staging}). In our example,
splitting transforms the function $f$ into two others $f_1$ and $f_2$
such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating the first function on $x$, and using the result $z$ to evaluate
the second function on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between splitting and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Therefore, unlike partial
evaluation, splitting is a static program transformation
(``metastatic'' in partial evaluation terms) and does not require
dynamic code generation.

@@

Prior work on partial evaluation and metaprogramming has demonstrated
automatic application of these techniques on higher order functional
languages. In contrast, automatic splitting transforms have been
limited to simple languages\,\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}.

@@

We demonstrate that splitting a recursive mixed-stage $f$
yields an $f_1$ which computes a recursive data structure
and an $f_2$ which traverses that structure
in light of information available at the second stage.
In the case of the quickselect algorithm, which we discuss next, 
the split code executes asymptotically faster than a unstaged evaluation of $f$.

>>>>

\input{figures/quickselect}
Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_1, y_2 \ldots, y_m)$@ = 
  @$f(x, y_1);  f(x, y_2); \ldots; f(x, y_m)$@
\end{lstlisting}
%
In this scenario, a multivariate function is applied many times with
one argument fixed, so it is advantageous to find the computations in
$f$ that depend on $x$ and then stage program execution so they are
performed only once at the beginning of execution.

@@

J{\o}rring and Scherlis classify automatic program staging
transformations, such as the one described above, as forms of {\em
  frequency reduction} or {\em
  precomputation}\,\cite{JS86-staging}. To perform frequency
reduction, one identifies and hoists computations that are performed
multiple times, in order to compute them only once. To perform
precomputation, one identifies computations that can be performed in
advance and does so---for example, at compile time if the relevant
inputs are statically known.

@@

While simple forms of frequency reduction include standard compiler
optimizations such as loop hoisting and common subexpression
elimination, J{\o}rring and Scherlis proposed the more general
transformation of {\em splitting} a program into multiple subfunctions
(called pass separation in \,\cite{JS86-staging}). In our example,
splitting transforms the function $f$ into two others $f_1$ and $f_2$
such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating $f_1$ on $x$, and using the result $z$ to evaluate
the second function on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between splitting and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Therefore, unlike partial
evaluation, splitting is a static program transformation
(``metastatic'' in partial evaluation terms) and does not require
dynamic code generation.

@@

Prior work on partial evaluation and metaprogramming has demonstrated
automatic application of these techniques on higher order functional
languages. In contrast, automatic splitting transformations have been
limited to simpler languages\,\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}.

@@

We also demonstrate that splitting a recursive mixed-stage $f$
yields an $f_1$ which computes a recursive data structure
and an $f_2$ which traverses that structure
in light of information available at the second stage.
In the case of the quickselect algorithm, which we discuss next, 
the split code executes asymptotically faster than an unstaged evaluation of $f$.

### 2015-07-11 paper/overview.tex
figure fiddling
various small edits

\input{figures/quickselect}

@@

A first cut at a solution might be to, in a first pass, sort \texttt{l} into
\texttt{s} and then in a second pass, perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since
\texttt{s} is a list, leading to no improvement in efficiency.

A more sophisticated solution is to, in a first pass, construct a binary search
tree out of \texttt{l} and in a second pass, simply look for the \texttt{k}th
leftmost element of that tree. To do this efficiently requires one more
innovation---augmenting the tree by storing at each node the size of its left
subtree. This approach has (expected) runtime $\Theta(n\log{n} + m\log{n})$, a
nearly linear-time improvement in $m$.

In both cases, we replace our simple solution with a two-pass algorithm which
first processes the input into a lookup data structure, and then subsequently
performs fast lookups. Improving algorithms in this way is highly non-trivial,
as it requires reasoning about intricate algorithmic concerns, and implementing
more complex algorithms and data structures.

In this paper, we present a splitting algorithm which automatically transforms
staged programs into multi-pass programs. In certain cases, this algorithm even
improves input algorithms in the way described above!

%Such transformations, called pass-separation by J{\o}rring and
%Scherlis are commonly employed.  For example, as briefly mentioned in
%\ref{sec:intro}, modern graphics software is written exactly in this
%way.  Unfortunately, as the example illustrates, multipass programs
%are naturally complex.

@@

The idea behind staged programming is to use staging annotations---in our case,
guided by types---to indicate the stage of each subterm. We consider a
language \lang\ with two modal worlds representing the two stages, and a third
{\em ground} world of unstaged terms.

@@

The process of automatically adding staging annotations to unstaged code has been
the subject of extensive research (\ref{sec:related}). In this paper, we do not
consider this problem, instead assuming that the annotations already exist.
In the case of \texttt{qss}, we have specifically chosen annotations which
maximize the work performed in the first stage.

@@

In the rest of this section, we present a high-level overview of the main ideas
behind our splitting algorithm, applied to \texttt{qss}.
%
Splitting \texttt{qss} yields a two-pass program that creates a
probabilistically balanced, augmented binary search tree (which is isomorphic to
a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
particular, its first pass (\texttt{qs1}) constructs such a binary search tree
and its second pass (\texttt{qs2}) traverses the tree, using the embedded size
information to find the element of the desired rank.
%
The code for \texttt{qs1} and \texttt{qs2} is in \ref{fig:qs-split}.

@@

Our splitting algorithm scans \texttt{qss} for first-stage computations,
gathering them into \texttt{qs1}. Given \texttt{l}, this pass performs these
computations and places the information needed by the subsequent pass into a
boundary data structure.
%
In particular, \texttt{qs1} performs all recursive calls and evaluates all
instances of \texttt{partition} (since it depends only on \texttt{l}). It
produces a boundary data structure that collects the results from these
recursive calls, tagged by the branch (\texttt{LT}, \texttt{EQ}, or \texttt{GT})
in which that call occurred. Since the recursive calls occur in two different
branches (\texttt{LT} and \texttt{GT}) the boundary structure is a binary tree.  
%
Lastly, it records \texttt{i} (the size of the left subtree) and \texttt{\#1 ht}
(the pivot/head of the list) in the boundary structure, because those
computations are \emph{held} for use in the second stage.
%
The final result is a binary search tree augmented with size information, and
whose keys are the pivots.

Our splitting algorithm simultaneously scans \texttt{qss} for second-stage
computations, gathering them into \texttt{qs2}. This pass is given the boundary
data structure and the rank \texttt{k}, and it finishes the computation. Now that
\texttt{k} is known, the conditional on \texttt{compare k i} can be evaluated,
choosing which recursive call of \texttt{qss} is actually relevant for this
\texttt{k}. Since the boundary data structure contains the first-stage data for
\emph{all} of the recursive calls, performing these comparisons essentially
walks the tree, using the rank along with the size data \texttt{i} in order to
look up the \texttt{k}th leftmost node in the tree.

>>>>

@@

A first cut at a solution might be to first sort \texttt{l} into
\texttt{s} and then perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since
\texttt{s} is a list, leading to no improvement in efficiency.

A more sophisticated solution is to construct a binary search
tree out of \texttt{l} and then simply look for the \texttt{k}th
leftmost element of that tree. To do this efficiently requires one more
innovation---augmenting the tree by storing at each node the size of its left
subtree. This approach has (expected) runtime $\Theta(n\log{n} + m\log{n})$, a
nearly linear-time improvement in $m$.

In both cases, we replace our simple solution with a two-part algorithm which
first processes the input into a lookup data structure, and then subsequently
performs fast lookups. Improving algorithms in this way is highly non-trivial,
as it requires reasoning about intricate algorithmic concerns, and implementing
more complex algorithms and data structures.

In this paper, we present a splitting algorithm which automatically transforms
staged programs into multi-part programs. In certain cases, this algorithm even
improves input algorithms in the way described above!

@@

The idea behind staged programming is to use staging annotations---in our case,
guided by types---to indicate the stage of each subterm.

@@

The process of automatically adding staging annotations to unstaged code,
called \emph{binding time analysis}, has been
the subject of extensive research (\ref{sec:related}). In this paper, we do not
consider this problem, instead assuming that the annotations already exist.
In the case of \texttt{qss}, we have specifically chosen annotations which
maximize the work performed in the first stage.

@@

In the rest of this section, we present a high-level overview of the main ideas
behind our splitting algorithm, applied to \texttt{qss}.
%
Splitting \texttt{qss} yields a two-part program that creates a
probabilistically balanced, augmented binary search tree (which is isomorphic to
a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
particular, its first part (\texttt{qs1}) constructs such a binary search tree
and its second part (\texttt{qs2}) traverses the tree, using the embedded size
information to find the element of the desired rank.
%
The code for \texttt{qs1} and \texttt{qs2} is in \ref{fig:qs-split}.

@@

Our splitting algorithm scans \texttt{qss} for first-stage computations,
gathering them into \texttt{qs1}. Given \texttt{l}, this function performs these
computations and places the information needed by the subsequent function into a
boundary data structure.
%
In particular, \texttt{qs1} performs all recursive calls and evaluates all
instances of \texttt{partition} (since it depends only on \texttt{l}). It
produces a boundary data structure that collects the results from these
recursive calls, tagged by the branch (\texttt{LT}, \texttt{EQ}, or \texttt{GT})
in which that call occurred. Since the recursive calls occur in two different
branches (\texttt{LT} and \texttt{GT}) the boundary structure is a binary tree.  
%
Lastly, it records \texttt{i} (the size of the left subtree) and \texttt{\#1 ht}
(the pivot/head of the list) in the boundary structure, because those
computations are \emph{held} for use in the second stage.
%
The final result is a binary search tree augmented with size information, and
whose keys are the pivots.

Our splitting algorithm simultaneously scans \texttt{qss} for second-stage
computations, gathering them into \texttt{qs2}. This function is given the boundary
data structure and the rank \texttt{k}, and it finishes the computation. Now that
\texttt{k} is known, the conditional on \texttt{compare k i} can be evaluated,
choosing which recursive call of \texttt{qss} is actually relevant for this
\texttt{k}. Since the boundary data structure contains the first-stage data for
\emph{all} of the recursive calls, performing these comparisons essentially
walks the tree, using the rank along with the size data \texttt{i} in order to
look up the \texttt{k}th leftmost node in the tree.

### 2015-07-11 paper/semantics.tex
various small edits

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input top-level program $\coltwo e A$
and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since a residual $q$ is monostage by definition, this second pass is standard unstaged evaluation
and is not described in further detail in this paper.
Moreover, for the purposes of these dynamics, we consider a top level program to always be typed at world \bbtwo.

Since $\coltwo e A$ may be constructed out of terms at other worlds, 
our dynamics requires notions of values and steps that are specialized to each world.

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement \mbox{$\steptwo e {e'}$}.
Since first pass evaluation should not reduce stage two terms, this judgement does nothing but traverse $e$ to find \prev\ blocks, under which it performs in-place reductions.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with an eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

@@

The statics and dynamics of \lang\ are related by the type safety theorems
in \ref{fig:typeSafety}, again annotated by world.
In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.

>>>>

The central tenet of a staged language
is that first-stage code should be evaluated entirely before second-stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input top-level program $\coltwo e A$
and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since $q$ is monostage by definition, this second pass is standard unstaged evaluation
and is not described in further detail in this paper.
Moreover, for the purposes of these dynamics, we consider a top level program to always be typed at world \bbtwo.

Since $\coltwo e A$ may be constructed out of terms at other worlds, 
our dynamics requires notions of values and steps that are specialized to each world.
The rules for all parts of first-pass evaluation are given in \ref{fig:diaSemantics}.
In this and later figures, we extensively use an $\mathcal{S}[\text{-}]$ construction
to indicate a \emph{shallow} evaluation context which looks a single level deep.

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgment \mbox{$\steptwo e {e'}$}.
Since first pass evaluation should not reduce stage two terms, this judgment does nothing but traverse $e$ to find \prev\ blocks, under which it performs in-place reductions.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with an eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgment, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

@@

The statics and dynamics of \lang\ are related by the type safety theorems
in \ref{fig:typeSafety}, again annotated by world.
In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.
Note how the progress theorem for world \bbonem\ states that every well-typed term must take 
either a standard step or a hoisting step.
